Gaussian Processes Autoregressive Models for the Disturbance Storm Time Index

### 2016-04-26 swsc.tex
Updates from ShareLaTeX

%Talk about Burton and friends
We focus on prediction of the hourly $D_{st}$ index which is an important problem because $D_{st}$ is a straight forward indicator of the occurrence of geomagnetic storms. More specifically we focus on the \emph{one step ahead} (OSA) prediction of $D_{st}$ because it is the first step towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

@@

%Talk about neural networks
Application of \emph{Machine Learning} methods for prediction of geomagnetic indices is fairly recent (1990's onwards). \emph{Neural Networks} have been a popular choice for building predictive models and researchers employed the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture for prediction of $D_{st}$. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which employed the \emph{Boyle coupling function} \citet{boyle1997empirical}, in \citet{SWE:SWE286} the same architecture is adapted for prediction of $D_{st}$ and $AE$ indices as well.

%Talk about comparison papers, number of storms covered and Persistence

>>>>

%Talk about Burton and friends
We focus on prediction of the hourly $D_{st}$ index which is an important problem because $D_{st}$ is a straight forward indicator of geomagnetic storms. More specifically we focus on the \emph{one step ahead} (OSA) prediction of $D_{st}$ because it is the first step towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

@@

%Talk about neural networks
\emph{Neural Networks} have been a popular choice for building predictive models and researchers employed the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $D_{st}$, which used the solar wind, \emph{interplanetary magnetic field} (IMF) and historical $D_{st}$ data. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} (\citet{boyle1997empirical}), in \citet{SWE:SWE286} the same architecture is adapted for prediction of $D_{st}$, which is popularly known as the Rice $D_{st}$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $D_{st}$ using only the IMF data.

%Talk about comparison papers, number of storms covered and Persistence
In light of the extensive list of modeling techniques employed for prediction of geomagnetic indices and in particular $D_{st}$, model comparison and evaluation becomes a crucial direction in advancing the research domain. \citet{Rastatter2013} compare several physics based (convection, kinetic and magneto hydrodynamic) and empirical prediction models such as Narmax, Rice $D_{st}$ on 4 storm events which occurred between 2001 and 2006. \citet{Amata2008} compares the EDDA and the Lund $D_{st}$ models over the 2003-2005 period. But the most extensive model comparison in terms of storm events was conducted in \citet{Ji2012} which compared seven $D_{st}$ models (see table \ref{table:DstModels}) on a list of 63 geomagnetic storm events of varying intensities which occurred between 1998 and 2006.

%Vision
In $D_{st}$ prediction a seemingly trivial yet highly informative prediction method known as the \emph{Persistence} model has been largely ignored in its crucial role in model evaluation. The \emph{Persistence} model is stated simply as using the previous value of $D_{st}$ as the prediction for the next time step ($\hat{D_{st}}(t) = D_{st}(t-1)$). Due to high auto correlation between $D_{st}(t)$ and $D_{st}(t-1)$ in the Omni data, the \emph{Persistence} model gives excellent predictive performance despite its apparent lack of sophistication. In essence it is a trivial hypothesis and hence must always be used as a base line to compare the performance of any proposed $D_{st}$ algorithm. Therefore in the context of OSA prediction, a model must outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction.

%Assertion
It is with the above vision in mind, we propose \emph{Gaussian Process} models for OSA prediction of $D_{st}$. We use the results of \citet{Ji2012} as a starting point and compare our proposed models with the methods outlined in the same paper while using the performance of the \emph{Persistence} model as a litmus test.
